---
title: "Full Analysis: Seasonality and Geography of Airport Flight Traffic"
author: "Kaden Zhu"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 4.5
)
```

```{r packages}
library(tidyverse)
library(scales)
library(broom)
library(knitr)
library(fixest)   # fast fixed-effects regression (install.packages("fixest"))
```

```{r helpers, include = FALSE}
# Tidy a coefficient table from either an lm (model1) or a fixest
# object (model2, model2_log), so the rest of the document treats them
# the same way.
tidy_coefs <- function(m) {
  if (inherits(m, "fixest")) {
    ct <- fixest::coeftable(m)
    tibble(
      term      = rownames(ct),
      estimate  = ct[, 1],
      std.error = ct[, 2],
      statistic = ct[, 3],
      p.value   = ct[, 4]
    )
  } else {
    broom::tidy(m)
  }
}

# Overall R-squared for either object type.
get_r2 <- function(m) {
  if (inherits(m, "fixest")) unname(fixest::r2(m, type = "r2"))
  else broom::glance(m)$r.squared
}
```

# 1. Load and combine the data

```{r load}
airport_2025 <- read_csv("data/airport_traffic_2025.csv")
airport_2026 <- read_csv("data/airport_traffic_2026.csv")

airport <- bind_rows(airport_2025, airport_2026) |>
  mutate(
    MONTH_MON = fct_reorder(MONTH_MON, MONTH_NUM),
    YEAR = factor(YEAR)
  )

dim(airport)
```

The combined frame holds every airport-day record from both years
side by side. Converting `MONTH_MON` to an ordered factor keeps the
calendar order on every plot and in the regression. Treating `YEAR`
as a factor with two levels matches what the linear model needs.

# 2. Data quality check

```{r missingness}
airport |>
  summarize(across(everything(), ~ sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  arrange(desc(n_missing))
```

```{r counts}
tibble(
  n_records = nrow(airport),
  n_airports = n_distinct(airport$APT_ICAO),
  n_countries = n_distinct(airport$STATE_NAME),
  n_dates = n_distinct(airport$FLT_DATE)
)
```

I read missingness first because anything substantial in `FLT_TOT_1`
would change the interpretation of every summary statistic below.
The record counts confirm the dataset's scale: hundreds of airports
across a couple of years gives us a sample large enough to fit a
regression with airport-level fixed effects without running out of
degrees of freedom.

# 3. Exploratory data analysis

## 3.1 Outcome distribution

```{r dist-hist}
ggplot(airport, aes(x = FLT_TOT_1)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Daily flight totals: heavily right-skewed",
    x = "Daily flight total (FLT_TOT_1)",
    y = "Number of airport-days"
  ) +
  theme_minimal(base_size = 12)
```

```{r dist-log}
ggplot(airport, aes(x = FLT_TOT_1 + 1)) +
  geom_histogram(bins = 60, fill = "darkorange", color = "white") +
  scale_x_log10(labels = comma) +
  labs(
    title = "Same distribution on a log10 scale",
    subtitle = "Approximately symmetric, a useful sign for residual diagnostics",
    x = "Daily flight total (log10 scale)",
    y = "Number of airport-days"
  ) +
  theme_minimal(base_size = 12)
```

The raw histogram has the long right tail I expected from the
boxplots in the proposal. On a log scale the distribution looks
roughly symmetric, which tells me a log transformation of the
outcome is a sensible thing to keep in reserve in case the regression
residuals misbehave.

## 3.2 Overall and grouped summaries

```{r summary-overall}
airport |>
  summarize(
    n = n(),
    mean = mean(FLT_TOT_1, na.rm = TRUE),
    median = median(FLT_TOT_1, na.rm = TRUE),
    sd = sd(FLT_TOT_1, na.rm = TRUE),
    p10 = quantile(FLT_TOT_1, 0.10, na.rm = TRUE),
    p90 = quantile(FLT_TOT_1, 0.90, na.rm = TRUE),
    max = max(FLT_TOT_1, na.rm = TRUE)
  ) |>
  kable(digits = 1, caption = "Overall summary of daily flight totals.")
```

The mean is well above the median and the 90th percentile is several
times the median. This is the right tail in numbers rather than in a
picture: the typical airport-day in this sample is much quieter than
the average suggests.

```{r summary-month}
airport |>
  group_by(MONTH_MON) |>
  summarize(
    n = n(),
    mean = mean(FLT_TOT_1, na.rm = TRUE),
    median = median(FLT_TOT_1, na.rm = TRUE),
    sd = sd(FLT_TOT_1, na.rm = TRUE),
    .groups = "drop"
  ) |>
  kable(digits = 1, caption = "Daily flight totals by month.")
```

Across both years the monthly means rise from a winter low through
spring, crest in mid-summer, and fall back through autumn. The medians track the means too. The seasonal pattern is not an
artifact of the right tail.

```{r summary-country}
top_countries <- airport |>
  group_by(STATE_NAME) |>
  summarize(mean_flights = mean(FLT_TOT_1, na.rm = TRUE), .groups = "drop") |>
  slice_max(mean_flights, n = 10) |>
  pull(STATE_NAME)

airport |>
  filter(STATE_NAME %in% top_countries) |>
  group_by(STATE_NAME) |>
  summarize(
    n = n(),
    n_airports = n_distinct(APT_ICAO),
    mean = mean(FLT_TOT_1, na.rm = TRUE),
    median = median(FLT_TOT_1, na.rm = TRUE),
    sd = sd(FLT_TOT_1, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(mean)) |>
  kable(digits = 1, caption = "Daily flight totals for the top 10 countries by mean.")
```

The `n_airports` column is the key one in this table. The countries
at the top of the ranking are not necessarily countries with many
busy airports. Some have only one or two airports in the sample, and
those airports happen to be very busy. That is exactly the confound
the regression is built to address.

## 3.3 Seasonal pattern

```{r plot-month}
airport |>
  group_by(MONTH_MON, YEAR) |>
  summarize(mean_flights = mean(FLT_TOT_1, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = MONTH_MON, y = mean_flights,
             group = YEAR, color = YEAR)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Average daily flight volume by month",
    subtitle = "Same seasonal shape in both calendar years",
    x = "Month",
    y = "Mean daily flights per airport",
    color = "Year"
  ) +
  theme_minimal(base_size = 12)
```

The two yearly lines move together across the calendar, so the
seasonal effect is reproducible rather than year-specific. Where the
lines diverge by a few flights per airport, that is the size of the
year effect we will recover in the regression.

## 3.4 Country comparison

```{r plot-country}
airport |>
  filter(STATE_NAME %in% top_countries) |>
  mutate(STATE_NAME = fct_reorder(STATE_NAME, FLT_TOT_1, .fun = median)) |>
  ggplot(aes(x = STATE_NAME, y = FLT_TOT_1)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6, outlier.alpha = 0.1) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Daily flight totals by country (top 10 by mean)",
    subtitle = "Wide spreads come from hub-vs-regional variation within countries",
    x = NULL,
    y = "Daily flight total"
  ) +
  theme_minimal(base_size = 12)
```

The width of each box is a within-country story. Countries with both
a major hub and several regional airports, such as Germany, the UK, and Italy,
get the widest spreads because the same country produces both
1,500-flight days at the hub and 50-flight days at small airports.

## 3.5 The airport-level confound made visible

```{r plot-airports}
airport |>
  group_by(APT_ICAO, STATE_NAME) |>
  summarize(mean_flights = mean(FLT_TOT_1, na.rm = TRUE), .groups = "drop") |>
  slice_max(mean_flights, n = 15) |>
  mutate(APT_ICAO = fct_reorder(APT_ICAO, mean_flights)) |>
  ggplot(aes(x = APT_ICAO, y = mean_flights, fill = STATE_NAME)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(
    title = "The 15 busiest airports dominate the right tail",
    subtitle = "A handful of hubs across a handful of countries",
    x = "Airport (ICAO)",
    y = "Mean daily flight total",
    fill = "Country"
  ) +
  theme_minimal(base_size = 12)
```

A small set of hubs sits far above the rest. When the boxplots above
made a country look busy, this plot is usually the reason. The
regression in section 4 puts these airports in as fixed effects so
that month and year coefficients are estimated after holding airport
identity constant.

# 4. Regression models

## 4.1 Model 1, without airport control

```{r model1}
model1 <- lm(FLT_TOT_1 ~ MONTH_MON + STATE_NAME + YEAR, data = airport)

glance(model1) |>
  select(r.squared, adj.r.squared, sigma, statistic, p.value, df, nobs) |>
  kable(digits = 3, caption = "Model 1 fit statistics.")
```

```{r model1-coefs}
tidy(model1) |>
  filter(str_starts(term, "MONTH_MON") | term == "YEAR2026") |>
  kable(digits = 2, caption = "Model 1: month and year coefficients (intercept = January, baseline year 2025).")
```

Model 1 is the naive picture. Its R² answers "how much of the
variation in daily flights can month, country, and year jointly
explain?" Its month coefficients give the average shift in daily
flights for each month relative to January, ignoring the fact that
the country mix doing those summer flights skews toward hub-heavy
states.

## 4.2 Model 2, with airport fixed effects

```{r model2}
# Absorb airport as a fixed effect instead of building 330 dummy
# columns, so the model fits quickly on the full data.
model2 <- feols(FLT_TOT_1 ~ MONTH_MON + YEAR | APT_ICAO, data = airport)

tibble(
  r.squared        = get_r2(model2),
  within.r.squared = fixest::r2(model2, type = "wr2"),
  n_airports_fe    = model2$fixef_sizes[["APT_ICAO"]],
  nobs             = nobs(model2)
) |>
  kable(digits = 3, caption = "Model 2 fit statistics (airport absorbed as fixed effect).")
```

```{r model2-coefs}
tidy_coefs(model2) |>
  filter(str_starts(term, "MONTH_MON") | term == "YEAR2026") |>
  kable(digits = 2, caption = "Model 2: month and year coefficients after controlling for airport.")
```

Model 2 is the model I actually want to interpret. Absorbing airport
as a fixed effect soaks up the "which hubs are in the sample" effect,
so the month coefficients here estimate the average within-airport
seasonal effect: how much busier (or quieter) an airport is in a
given month compared to its own January baseline. `STATE_NAME` is
not in this model. Country is fully determined by airport, so
including both would be perfectly collinear.

## 4.3 Comparing the two models

```{r compare}
bind_rows(
  tidy_coefs(model1) |> mutate(model = "Model 1 (no airport control)"),
  tidy_coefs(model2) |> mutate(model = "Model 2 (airport fixed effects)")
) |>
  filter(str_starts(term, "MONTH_MON")) |>
  mutate(month = str_remove(term, "MONTH_MON"),
         month = fct_reorder(month, match(month, levels(airport$MONTH_MON)))) |>
  ggplot(aes(x = month, y = estimate, color = model, group = model)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = estimate - 1.96 * std.error,
                    ymax = estimate + 1.96 * std.error),
                width = 0.2, alpha = 0.6) +
  labs(
    title = "Month coefficients are stable across both models",
    subtitle = "Reference month is January; bars are 95% confidence intervals",
    x = "Month", y = "Estimated change in daily flights vs January",
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

If the seasonal pattern were really a story about hub composition,
the teal line (Model 2) would flatten. The closeness of the two
lines shows the monthly effect is within-airport seasonal variation, not the byproduct of which
countries happen to have records that month.

## 4.4 Diagnostics

```{r diagnostics, fig.width = 8, fig.height = 4, dev = "png", dpi = 150}
# plot(model2) would compute leverage for every observation, slow at
# this size. Residuals and fitted values are already stored, so sample
# 5000 points from those for the diagnostic plots.
set.seed(1)
idx <- sample(seq_len(nrow(airport)), min(5000, nrow(airport)))
fit <- fitted(model2)[idx]
res <- resid(model2)[idx]

par(mfrow = c(1, 2))
plot(fit, res, pch = ".", col = rgb(0, 0, 0, 0.3),
     xlab = "Fitted", ylab = "Residual", main = "Residuals vs fitted")
abline(h = 0, col = "red")
qqnorm(res, pch = ".", main = "Normal Q-Q")
qqline(res, col = "red")
par(mfrow = c(1, 1))
```

The residual-vs-fitted plot is where the right skew shows up: if the
spread of residuals fans out as fitted values grow, that is
heteroskedasticity, and it is the reason I refit on the log scale
below and report those coefficients as percent changes. The Q-Q plot
checks whether the residuals are roughly normal. Heavy tails there
point in the same direction.

## 4.5 Log-scale robustness check

```{r model2-log}
model2_log <- feols(log(FLT_TOT_1 + 1) ~ MONTH_MON + YEAR | APT_ICAO, data = airport)

tibble(
  r.squared        = get_r2(model2_log),
  within.r.squared = fixest::r2(model2_log, type = "wr2"),
  nobs             = nobs(model2_log)
) |>
  kable(digits = 3,
        caption = "Model 2 refit on log(FLT_TOT_1 + 1). Coefficients are approximate proportional changes.")

tidy_coefs(model2_log) |>
  filter(str_starts(term, "MONTH_MON") | term == "YEAR2026") |>
  mutate(pct_change = (exp(estimate) - 1) * 100) |>
  kable(digits = 3, caption = "Log-scale month and year coefficients, with implied percent change.")
```

On the log scale a coefficient of 0.10 for July means daily traffic
at an average airport runs roughly 10% higher in July than in January.
I report this version in the executive summary and slides. The log
form drops the assumption of equal additive effects across airports
of widely different sizes.

# 5. Interpretation in domain terms

The two models point to one conclusion. Daily flight volume rises and
falls with the season. Summer months sit well above the January
baseline at the average airport. The country mix in the sample does
not drive this shape. Month coefficients barely move when airport
fixed effects enter the model. July adds about 52 flights in Model 1
and about 54 in Model 2, near-identical figures.

The country effect behaves the opposite way. Model 1 ranks Denmark
and Finland near the top. The country table explains why. Denmark
contributes one airport, Copenhagen. Finland contributes one airport,
Helsinki. Their high averages describe two busy hubs, not two busy
countries. Once airport enters the model, those country signals fold
into the airport fixed effects and lose meaning.

The year coefficient compares 2026 to 2025 with month and airport held
constant. The effect is small and sensitive to scale. In the level
model, 2026 runs about 2 flights per day above 2025. In the log model,
2026 runs about 2 percent below. The two specifications split on sign
and agree on size, close to zero. In practical terms, 2025 and 2026
look alike.

# 6. Limitations

Four limitations shape the conclusions.

The geographic scope is narrow. The source is EUROCONTROL coverage.
The findings describe European IFR traffic, not global aviation. North
American, East Asian, and African airports sit outside the sample.

The outcome mixes flight types. Daily total movements combine
passenger, cargo, business, and general aviation. A 200-movement day
at a cargo hub and a 200-movement day at a tourist airport count the
same. Separating them needs data the EUROCONTROL daily extract does
not provide.

Two years limits the seasonal claim. The pattern repeats across 2025
and 2026, which rules out a one-year fluke. Two years still does not
separate season from year-specific shocks. A strike, a volcanic
closure, or a geopolitical event in one summer would enter the model
as a month or year effect with no warning sign.

The design is observational. Airport fixed effects remove
airport-level confounding, not all confounding. The data supports one
claim. July runs busier than January at the average airport. The data
does not support a causal claim about summer demand.

# 7. What I would do next

Three extensions would strengthen the work.

Split the outcome into departures and arrivals. This would show
whether tourism creates timing asymmetries, more arrivals in early
summer and more departures in late summer at sun destinations.

Add country-level predictors. GDP per capita, tourist-arrival share,
and public-holiday calendars would give the country comparison real
content. The current model drops country because airport absorbs the
country signal, so these predictors would need a design without
airport fixed effects.

Fit a mixed-effects model with random airport intercepts. This would
return the same airport-controlled month estimates and report how much
unexplained variance sits at the airport level versus the day level.
