---
title: "Seasonality and Geography of Airport Flight Traffic"
author: "by Kaden Zhu"
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(broom)
library(fixest)   # install.packages("fixest")
```

```{r load-data, echo = FALSE, message = FALSE}
airport <- bind_rows(
  read_csv("data/airport_traffic_2025.csv"),
  read_csv("data/airport_traffic_2026.csv")
) |>
  mutate(
    MONTH_MON = fct_reorder(MONTH_MON, MONTH_NUM),
    YEAR = factor(YEAR)
  )

median_daily <- round(median(airport$FLT_TOT_1, na.rm = TRUE), 0)
mean_daily   <- round(mean(airport$FLT_TOT_1, na.rm = TRUE), 0)
n_apts       <- n_distinct(airport$APT_ICAO)
n_states     <- n_distinct(airport$STATE_NAME)

# Model 1 is small (country has ~30 levels), so plain lm is fine.
# Models 2 use feols to absorb the 331-level airport factor efficiently.
model1     <- lm(FLT_TOT_1 ~ MONTH_MON + STATE_NAME + YEAR, data = airport)
model2     <- feols(FLT_TOT_1 ~ MONTH_MON + YEAR | APT_ICAO, data = airport)
model2_log <- feols(log(FLT_TOT_1 + 1) ~ MONTH_MON + YEAR | APT_ICAO, data = airport)

m1_r2 <- round(glance(model1)$r.squared, 2)
m2_r2 <- round(unname(fixest::r2(model2, type = "r2")), 2)
jul_pct <- round((exp(coef(model2_log)[["MONTH_MONJUL"]]) - 1) * 100, 0)
```

## Summary

This project asks whether daily flight totals at international
airports depend on the time of year and the country an airport sits
in, or whether the apparent role of those factors comes down to which
airports the data happens to include. Seasonality and country mix are exactly the levers that
airport planners, tourism boards, and aviation-emissions analysts
reach for, so getting their relative importance right matters.
Country effects in aviation data often turn out to be hub-composition
effects rather than real country differences. I built the analysis
from the start to separate the two.

The data is a Kaggle release of EUROCONTROL-style airport traffic
statistics: one row per airport per day, recording total
instrument-flight-rules movements, across the 2025 and 2026 calendar
years. After combining the two yearly files the sample covers
`r n_apts` airports across `r n_states` countries, weighted heavily
toward Europe with a few neighboring states. The outcome variable is
`FLT_TOT_1`, daily total movements. The predictors of interest are
month and country, with airport ICAO code and calendar year as
controls.

Exploratory work came first. A histogram of the outcome showed a
strong right skew: the typical airport-day runs around `r median_daily`
movements, but the mean is `r mean_daily`, pulled upward by a handful
of mega-hubs running well over a thousand a day. A month-by-month line
plot, drawn separately for each year, showed a clear and reproducible
seasonal arc: traffic climbs from a winter low through spring, crests
in mid-summer, and falls back through autumn, with the two years
tracing almost identical shapes. A bar chart of the fifteen busiest
airports made the central confound visible before any model was fit.
A small set of hubs sits far above everything else, and each one is
the reason its country looks busy in a country-level comparison.

The modeling stage used two linear regressions on purpose, side by
side. The first regressed daily flights on month, country, and year
with no airport control. The second swapped country out for
airport-level fixed effects, since every airport sits in exactly one
country and including both would be perfectly collinear. The contrast
between the two is the whole point. If the seasonal effect is real and
operates within airports, its month coefficients should barely move
when airport fixed effects are introduced. If the country differences
are mostly a hub-composition story, their apparent size in the first
model should be absorbed by the airport dummies in the second. A
log-transformed version of the second model served as a robustness
check against the right tail in the outcome.

The results came out cleanly on both counts. The month coefficients
were nearly identical across the two models, which is strong
evidence that seasonality is a real within-airport effect rather
than an artifact of which countries had records in which months. In
the log-scale model, July traffic at the typical airport runs roughly
`r jul_pct`% above its own January baseline, holding airport and year
fixed, and the effect is highly significant given the large
sample. The country story went the opposite way, exactly as
predicted: countries that topped the raw rankings turned out to be
single-hub stories, with the "country effect" dissolving once airport
identity entered the model. The R² rose from about `r m1_r2` in the
first model to `r m2_r2` in the second, and almost all of that jump is
the airport fixed effects. Most of the variation in daily flights
sits between airports. Seasonality is a real but smaller effect on
top of each airport's own baseline.

Four limitations bound these conclusions. The geographic scope is
mostly European, so the findings describe EUROCONTROL airspace
rather than global aviation. The outcome lumps passenger, cargo,
business, and general aviation into one daily count, so the analysis
cannot speak to the mode mix behind the seasonal pattern even though
the summer peak strongly suggests leisure travel. Two calendar years
confirm a stable seasonal pattern but cannot separate seasonality from
one-off events such as strikes or weather disruptions. And the data is
observational, so even with airport fixed effects the right verb is
"associated with," not "caused by." A natural next step would be
splitting the outcome into arrivals and departures and adding
country-level covariates like tourism share and holiday calendars,
which the current model cannot use because airport absorbs country.

## Presentation

Presentation can be found [here](presentation/presentation.html).

## Data

European Organisation for the Safety of Air Navigation (EUROCONTROL).
*Airport Traffic Daily Movements, 2025-2026* [data set]. Distributed
via Kaggle. Retrieved [add your retrieval date], from
[add the Kaggle dataset URL you downloaded from].

## References

- European Organisation for the Safety of Air Navigation
  (EUROCONTROL). *Airport Traffic* statistics. [add Kaggle URL],
  retrieved [date].
- Wickham, H., et al. (2019). Welcome to the tidyverse. *Journal of
  Open Source Software*, 4(43), 1686.
- Robinson, D., Hayes, A., & Couch, S. (2023). *broom: Convert
  Statistical Objects into Tidy Tibbles.* R package.
