---
title: "Project proposal"
author: "Kaden Zhu"
output: github_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(scales)
library(skimr)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 4.5
)
```

```{r load-data, message = FALSE}
airport_2025 <- read_csv("../data/airport_traffic_2025.csv")
airport_2026 <- read_csv("../data/airport_traffic_2026.csv")

airport <- bind_rows(airport_2025, airport_2026)
```

## 1. Introduction

This project asks whether month of year and country of origin are
associated with daily flight traffic at international airports, and
whether those associations survive once we account for which specific
airports are in the data. The dataset compiles daily arrival and
departure counts for 331 airports across 2025 and 2026. Coverage is
heaviest in Europe but stretches to a handful of neighboring states
including Türkiye and Israel. The records come from a Kaggle release
of EUROCONTROL airport traffic statistics, which are official
instrument-flight-rules counts rather than estimates or scraped
figures.

The question matters because seasonality and country mix are the two
levers most often invoked when airports plan staffing, when tourism
boards model arrivals, and when emissions analysts build aviation
inventories. If summer really is a uniform European peak, that
supports one kind of planning. If "summer" is mostly a story about a
handful of Mediterranean hubs while northern Europe stays flat, that
is a very different story even though the marginal monthly average
might look the same.

Each row of the dataset is a single airport-day. The response variable
is `FLT_TOT_1`, the total IFR flight movements at that airport on that
date. The predictors of interest are `MONTH_MON` (month abbreviation)
and `STATE_NAME` (country). `APT_ICAO` (the airport itself) and `YEAR`
enter as controls.

## 2. Data

```{r glimpse-data}
glimpse(airport)
```

```{r skim-missing}
airport |>
  select(APT_ICAO, STATE_NAME, MONTH_MON, YEAR, FLT_TOT_1) |>
  skim()
```

The `skim()` output confirms the variable types and gives a first read
on missingness in the columns I rely on. Anything above a fraction of
a percent missing in `FLT_TOT_1` would be worth investigating before
fitting the regression; otherwise I treat occasional gaps as
ignorable. A full data dictionary lives in `data/README.md`.

## 3. Data analysis plan

**Variables.** The outcome (Y) is `FLT_TOT_1`, daily total IFR flight
movements. The predictors of interest (X) are `MONTH_MON`, the
seasonal signal, and `STATE_NAME`, the geographic signal. `APT_ICAO`
(airport) and `YEAR` enter as controls. Including airport is what lets
me separate "month matters" from "the busy airports happen to be in
our sample."

**Comparison groups.** The natural comparisons are across the twelve
calendar months and across countries. I restrict country comparisons
to the top ten countries by mean traffic so group sizes are roughly
comparable and no inference is driven by a country with only one or
two airports in the sample.

### Preliminary exploratory analysis

**Overall summary of the outcome.**

```{r summary-overall}
airport |>
  summarize(
    n_records = n(),
    n_airports = n_distinct(APT_ICAO),
    n_countries = n_distinct(STATE_NAME),
    mean_daily = mean(FLT_TOT_1, na.rm = TRUE),
    median_daily = median(FLT_TOT_1, na.rm = TRUE),
    sd_daily = sd(FLT_TOT_1, na.rm = TRUE),
    min_daily = min(FLT_TOT_1, na.rm = TRUE),
    max_daily = max(FLT_TOT_1, na.rm = TRUE)
  )
```

The mean sits well above the median, and the maximum is many times the
mean. That tells me the outcome is right-skewed: a small group of
mega-hubs do many more movements per day than the typical airport in
the sample. This matters for modeling because plain linear regression
assumes roughly symmetric residuals, so I check the distribution
before committing to a transformation.

**Distribution of the outcome.**

```{r viz-outcome-distribution, fig.cap = "Daily flight totals across all airport-days. Heavy right tail."}
ggplot(airport, aes(x = FLT_TOT_1)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Daily flight totals are heavily right-skewed",
    x = "Daily flight total (FLT_TOT_1)",
    y = "Number of airport-days"
  ) +
  theme_minimal(base_size = 12)
```

Most airport-days sit in the low-to-mid hundreds of movements, with a
long tail running into the low thousands at the biggest hubs. I keep
`FLT_TOT_1` on its original scale for descriptive statistics so the
units stay intuitive, but I plan to consider a `log(FLT_TOT_1 + 1)`
transformation for the regression if residual diagnostics show the
skew is causing trouble.

**Summary by month.**

```{r summary-by-month}
airport |>
  group_by(MONTH_NUM, MONTH_MON) |>
  summarize(
    n = n(),
    mean_daily = mean(FLT_TOT_1, na.rm = TRUE),
    median_daily = median(FLT_TOT_1, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(MONTH_NUM)
```

The means climb from a winter low through spring and crest in
mid-summer, then fall through autumn back to a winter trough. The
medians track the means closely month by month, which is a useful
sign: the seasonal pattern is not just being driven by the heavy
right tail. If summer were only "big airports get bigger," the medians
would barely move.

**Seasonal line plot.**

```{r viz-monthly-trend, fig.cap = "Mean daily flights per airport by month, averaged across all airport-days."}
airport |>
  mutate(MONTH_MON = fct_reorder(MONTH_MON, MONTH_NUM)) |>
  group_by(MONTH_MON) |>
  summarize(mean_flights = mean(FLT_TOT_1, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = MONTH_MON, y = mean_flights, group = 1)) +
  geom_line(linewidth = 1, color = "steelblue") +
  geom_point(size = 2, color = "steelblue") +
  labs(
    title = "Average daily flight volume by month",
    subtitle = "Winter low, mid-summer peak",
    x = "Month", y = "Mean daily flights per airport"
  ) +
  theme_minimal(base_size = 12)
```

Flights rise from a January low near 120 per airport to a July and
August peak near 170, then fall through autumn. The data covers all
of 2025 plus January 2026, so this shape comes from the twelve months
of 2025.

**Summary by country (top 10 by mean traffic).**

```{r summary-by-country}
top_countries <- airport |>
  group_by(STATE_NAME) |>
  summarize(mean_flights = mean(FLT_TOT_1, na.rm = TRUE), .groups = "drop") |>
  slice_max(mean_flights, n = 10) |>
  pull(STATE_NAME)

airport |>
  filter(STATE_NAME %in% top_countries) |>
  group_by(STATE_NAME) |>
  summarize(
    n_records = n(),
    n_airports = n_distinct(APT_ICAO),
    mean_daily = mean(FLT_TOT_1, na.rm = TRUE),
    median_daily = median(FLT_TOT_1, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(mean_daily))
```

The `n_airports` column is the key one. Some countries near the top
have only one or two airports in the sample, so their "country effect"
is the identity of one big airport. This is the exact
reason the regression has to include airport.

**Boxplot by country.**

```{r viz-country-boxplot, fig.cap = "Daily flight totals for the 10 countries with the highest mean traffic."}
airport |>
  filter(STATE_NAME %in% top_countries) |>
  mutate(STATE_NAME = fct_reorder(STATE_NAME, FLT_TOT_1, .fun = median)) |>
  ggplot(aes(x = STATE_NAME, y = FLT_TOT_1)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6, outlier.alpha = 0.1) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Daily flight totals by country (top 10 by mean traffic)",
    subtitle = "Boxes show variation within each country's airports",
    x = NULL, y = "Daily flight total"
  ) +
  theme_minimal(base_size = 12)
```

The boxes are wide and the whiskers reach far for countries that host
both a major hub and several regional airports. The very low medians
next to long upper tails are the visual signature of mega-hub outliers
pulling the mean upward.

**Top 15 airports by mean daily traffic.**

```{r viz-top-airports, fig.cap = "The fifteen busiest airports drive the right tail and much of the apparent country variation."}
airport |>
  group_by(APT_ICAO, STATE_NAME) |>
  summarize(mean_flights = mean(FLT_TOT_1, na.rm = TRUE), .groups = "drop") |>
  slice_max(mean_flights, n = 15) |>
  mutate(APT_ICAO = fct_reorder(APT_ICAO, mean_flights)) |>
  ggplot(aes(x = APT_ICAO, y = mean_flights, fill = STATE_NAME)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(
    title = "The 15 busiest airports drive the right tail",
    subtitle = "Bars colored by country show how few states host the largest hubs",
    x = "Airport (ICAO code)", y = "Mean daily flight total", fill = "Country"
  ) +
  theme_minimal(base_size = 12)
```

This is the plot I most wanted before fitting the model. A small
number of hubs sit far above the rest, and each is the reason its
country looks busy on the boxplot. Including airport as a control is
what makes the country and month coefficients interpretable as
something other than "where the hubs live."

### Proposed methods

I plan to fit two linear regressions, deliberately, to make the
confound visible. The first, `FLT_TOT_1 ~ MONTH_MON + STATE_NAME +
factor(YEAR)`, lets country and month compete for explanatory power on
their own; its coefficients conflate "country effects" with "which big
airports are in our sample." The second, `FLT_TOT_1 ~ MONTH_MON +
APT_ICAO + factor(YEAR)`, uses airport-level fixed effects. Because
every airport sits in exactly one country, including `APT_ICAO` makes
`STATE_NAME` redundant, so the country effect is not merely shrunk but
unidentifiable, and I drop it from the second model on purpose. What
the second model gives me cleanly is the month coefficients holding
airport identity fixed. If residual diagnostics flag
heteroskedasticity from the right skew, I refit on `log(FLT_TOT_1 + 1)`
and reinterpret the coefficients as approximate percent changes.

### Hypothesized results

I expect month to be the dominant signal in the second model, with
summer months (June to August) showing meaningfully higher daily flight
totals than January, and a secondary autumn dip. I expect the apparent
country differences in the first model to be mostly absorbed by airport
in the second, indicating that what looks like a "country effect" in
the boxplot is largely a story about which hubs the sample contains.
Results consistent with that prediction would show significant positive
summer month coefficients that stay stable across both models, and a
much larger R² for the second model driven by airport rather than
country. If summer effects collapse once airport is controlled, the
seasonal signal is itself a hub story rather than a Europe-wide pattern.
