---
title: "Seasonality and Geography of Airport Flight Traffic"
subtitle: "Math 225 Final Project"
author: "[your name]"
institute: "University of Edinburgh"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r load-packages, include = FALSE}
# Add any additional packages you need to this chunk
library(tidyverse)
library(broom)
library(knitr)
library(xaringanthemer)
library(scales)
```

```{r setup, include=FALSE}
# For better figure resolution
knitr::opts_chunk$set(
  fig.retina = 3, dpi = 300,
  fig.width = 6, fig.asp = 0.618,
  out.width = "80%",
  echo = FALSE, warning = FALSE, message = FALSE
)
```

```{r load-data, include=FALSE}
# Load your data here
airport <- bind_rows(
  read_csv("../data/airport_traffic_2025.csv"),
  read_csv("../data/airport_traffic_2026.csv")
) |>
  mutate(
    MONTH_MON = fct_reorder(MONTH_MON, MONTH_NUM),
    YEAR = factor(YEAR)
  )

n_rec <- nrow(airport)
n_apts <- n_distinct(airport$APT_ICAO)
n_states <- n_distinct(airport$STATE_NAME)
mean_daily <- round(mean(airport$FLT_TOT_1, na.rm = TRUE), 0)
median_daily <- round(median(airport$FLT_TOT_1, na.rm = TRUE), 0)

top_countries <- airport |>
  group_by(STATE_NAME) |>
  summarize(mean_flights = mean(FLT_TOT_1, na.rm = TRUE), .groups = "drop") |>
  slice_max(mean_flights, n = 10) |>
  pull(STATE_NAME)

model1 <- lm(FLT_TOT_1 ~ MONTH_MON + STATE_NAME + YEAR, data = airport)
model2 <- lm(FLT_TOT_1 ~ MONTH_MON + APT_ICAO + YEAR, data = airport)
model2_log <- lm(log(FLT_TOT_1 + 1) ~ MONTH_MON + APT_ICAO + YEAR, data = airport)

m1_r2 <- round(glance(model1)$r.squared, 3)
m2_r2 <- round(glance(model2)$r.squared, 3)
jul_pct <- round((exp(tidy(model2_log) |> filter(term == "MONTH_MONJUL") |> pull(estimate)) - 1) * 100, 1)
aug_pct <- round((exp(tidy(model2_log) |> filter(term == "MONTH_MONAUG") |> pull(estimate)) - 1) * 100, 1)
```

```{r}
style_xaringan(
  title_slide_background_image = "img/confetti.jpg"
)
```


class: center, middle

## Does seasonality and country drive daily airport flight traffic, or is it really about which hubs are in the sample?

---

class: inverse, center, middle

# The question

---

# Two reasons it matters

Airports plan staffing, runway slots, and ground capacity around
expected seasonal demand.

--

"Country effects" in aviation data are notorious for being
hub-composition effects in disguise. If we cannot tell those apart,
every comparison across countries is contaminated.

--

So the question is built around a contrast: is the seasonal signal
real *within* airports, and how much of the apparent country effect
is really just *where the hubs live*?

---

class: inverse, center, middle

# The data

---

# Daily airport traffic, 2025–2026

.pull-left[
- Kaggle release of EUROCONTROL-style daily airport traffic
- `r format(n_rec, big.mark = ",")` airport-day records
- `r n_apts` airports across `r n_states` countries
- Two full calendar years
- Outcome: `FLT_TOT_1` — total IFR flight movements per day
- Predictors: month, country, airport ICAO, year
]

.pull-right[
Median airport-day:
### `r median_daily` movements

Mean airport-day:
### `r mean_daily` movements

The gap between them is the right tail of mega-hubs.
]

---

# The shape of the outcome

```{r slide-hist, out.width = "85%", fig.asp = 0.5}
ggplot(airport, aes(x = FLT_TOT_1)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(x = "Daily flight total", y = "Number of airport-days") +
  theme_minimal(base_size = 14)
```

Strong right skew. A few hubs are very busy, most airports are quiet.
This is why airport identity has to be a control.

---

class: inverse, center, middle

# Exploratory patterns

---

# The seasonal pattern

```{r slide-month, out.width = "85%", fig.asp = 0.5}
airport |>
  group_by(MONTH_MON, YEAR) |>
  summarize(mean_flights = mean(FLT_TOT_1, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = MONTH_MON, y = mean_flights, group = YEAR, color = YEAR)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(x = "Month", y = "Mean daily flights per airport", color = "Year") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```

Both years trace nearly the same shape — peaks in June–August,
trough in January–February.

---

# The confound

```{r slide-airports, out.width = "85%", fig.asp = 0.55}
airport |>
  group_by(APT_ICAO, STATE_NAME) |>
  summarize(mean_flights = mean(FLT_TOT_1, na.rm = TRUE), .groups = "drop") |>
  slice_max(mean_flights, n = 15) |>
  mutate(APT_ICAO = fct_reorder(APT_ICAO, mean_flights)) |>
  ggplot(aes(x = APT_ICAO, y = mean_flights, fill = STATE_NAME)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(x = "Airport (ICAO)", y = "Mean daily flights", fill = "Country") +
  theme_minimal(base_size = 12)
```

Fifteen hubs do most of the heavy lifting. Country averages are
largely a story about which of these airports a country hosts.

---

class: inverse, center, middle

# Modeling

---

# Two regressions, on purpose

**Model 1** (naive):
$$\text{FLT\_TOT\_1} \sim \text{month} + \text{country} + \text{year}$$

**Model 2** (controlled):
$$\text{FLT\_TOT\_1} \sim \text{month} + \text{airport} + \text{year}$$

--

Airport is nested in country, so model 2 absorbs country. The
contrast is the point:

- If month is real seasonality, its coefficients **stay stable**
  across both models.
- If country is a hub-composition story, country's apparent effect
  in model 1 **disappears** when airport enters in model 2.

---

# Month effects are stable

```{r slide-coefs, out.width = "85%", fig.asp = 0.5}
bind_rows(
  tidy(model1) |> mutate(model = "Model 1 (no airport control)"),
  tidy(model2) |> mutate(model = "Model 2 (airport fixed effects)")
) |>
  filter(str_starts(term, "MONTH_MON")) |>
  mutate(month = str_remove(term, "MONTH_MON"),
         month = fct_reorder(month, match(month, levels(airport$MONTH_MON)))) |>
  ggplot(aes(x = month, y = estimate, color = model, group = model)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  labs(x = "Month", y = "Change in daily flights vs January", color = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```

The two lines sit on top of each other. Seasonality is genuine, not
a hub-composition artifact.

---

# In plain numbers

On the log scale, holding airport and year fixed:

.pull-left[
### July
**`r jul_pct`%**
higher than January at the average airport
]

.pull-right[
### August  
**`r aug_pct`%**
higher than January at the average airport
]

--

R² jumps from **`r m1_r2`** in model 1 to **`r m2_r2`** in model 2.

Most of the increase is the airport fixed effects: most variation in
daily flights is *between* airports, not *across* months.
Seasonality is real but small in absolute terms compared to who
you are.

---

class: inverse, center, middle

# Limitations and conclusions

---

# Limitations

- **Geographic scope.** EUROCONTROL coverage — findings generalize to
  European IFR traffic, not to global aviation.

--

- **Aggregated outcome.** Daily totals do not separate passenger,
  cargo, business, and general aviation.

--

- **Two years.** Enough to confirm a stable seasonal pattern, not
  enough to disentangle seasonality from one-off shocks like
  strikes or weather.

--

- **Observational data.** Even with airport fixed effects, this
  describes associations rather than causes.

---

# Takeaways

1. Daily flight traffic is **seasonally driven**, peaking June–August
   at roughly `r jul_pct`% above January at the typical airport.

--

2. **Apparent country differences are largely a hub-composition
   story.** Once airport identity is controlled, the country signal
   is absorbed.

--

3. The right modeling choice was to use **airport as a fixed effect**
   rather than country + airport together — doing the latter would
   have produced perfectly collinear predictors.

---

class: center, middle, inverse

# Thank you

### Questions?
